---
title: "00_ScreenRegionSelection"
output: html_document
date: "2026-02-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(bedtoolsr)
library(tidyverse)
library(readr)
```

#Candidate enhancers
## Step1: Region must be located in the IFNG locus
I first downloaded the Yates et al, 2019 dataset. Peak universe was created with XYZ. The Peak Universe contained 284,037 peaks. The peak universe was first filtered to keep regions in the IFNG locus chr12:68061040–68687723, performed with bedtools.

```{r}
PeakU<-read.delim("SNIP-R_ScreenRegionSelection/Yates_all_human_Tcell_ATAC.hg19.merged.bed",header=F)
IFNGlocus<-read.delim("SNIP-R_ScreenRegionSelection/IFNGlocus.bed.txt",header=F)
step1<-bedtoolsr::bt.intersect(a=PeakU,b=IFNGlocus,wa=T)

```

##Step2: Region must have at least one differential accessible condition
To enrich for functional regulatory elements. We filtered for peaks with at least one comparison with differential accessibility across the conditions in Yates et al, 2019 supplementary 2. We found the least significant (largest) Q-value from each of the pairwise comparison in the supplementary 2 dataset. We created a bed file containing regions whose least significant Q-value was above cutoff of 0.05. The bed files containing the non-differential regions were removed with bedtools subtract code below in commandline:

```{r}
NonDiffQ<-read.delim("SNIP-R_ScreenRegionSelection/Yates_nondiffOCR_chr12.bed",header=F)
Step2<-bedtoolsr::bt.subtract(a=step1,b=NonDiffQ)
```
There are 134 regions.

##Step3: Region must be also accessible in vitro T cell culture conditions where screen will be performed

```{r}
Invitro<-read.delim("SNIP-R_ScreenRegionSelection/InVitro_OCR_chr12.bed",header=F)
Step3<-bedtoolsr::bt.intersect(a=Step2,b=Invitro,u=TRUE)
```
There are 62 regions.

##Step4: Region must not be a protein coding region and region must have enough space for gRNA design
###Step4A: remove if direct protein coding overlap
We will remove any OCRs that overlap with protein coding region from UCSC protein coding bed track (hg19).
```{r}
ProteinCodingGene<-read.delim("SNIP-R_ScreenRegionSelection/UCSC_proteincodinggenes_hg19.bed")
Step4A<-bedtoolsr::bt.subtract(a=Step3, b=ProteinCodingGene,A=TRUE)
```

###Step4B:adjust region if too close to protein coding gene
We will need to remove all regions without enough space to design gRNAs because they are way to close to a protein coding region. Guide design without perturbing protein coding gene is improbable. To do this, we will create a 500bp window around the protein coding genes using bedtools slop. In this pipeline, we give around 500bp window for designing sgRNAs for SNIP-R deletion.

####1) Create 500bp protein coding gene slop (PCG-slop)
```{r}
genome_file <- normalizePath("SNIP-R_ScreenRegionSelection/human.hg19.genome")
ProteinCodingGene_Slop500<-bedtoolsr::bt.slop(i=ProteinCodingGene,g=genome_file,b=500)
```

####2) Find overlaps with the slopped regions - there were 2 overlaps.
Type 1: If ROI/candidate enhancer is fully engulfed by PCG-slop, then the entire ROI/candidate enhancer was removed.

```{r}
#Identifying type 1 overlaps
FullOverlapwithSlop<-bedtoolsr::bt.intersect(a=Step4A,b=ProteinCodingGene_Slop500,u=TRUE,f=1)

#Remove type 1 overlaps
Step4B<-bedtoolsr::bt.subtract(a=Step4A,b=FullOverlapwithSlop,A=TRUE)
```

Type2: If ROI/candidate enhancer is partially overlapping with PCG-slop, then the ROI/candidate boudaries were adjusted. More specifically, the ROI/candidate boudnaries were guided by the corresponding peak from another ATAC-seq dataset of relevance. In this case, I replaced it with in vitro ATAC-seq peak as it is the relevant screening approach. I made sure that the new replacement peak was not causing overalp with PCG-slop.

```{r}
#Identification of type2 overlaps
PartialOverlapwithSlop<-bedtoolsr::bt.intersect(a=Step4B,b=ProteinCodingGene_Slop500,u=TRUE)

#Get the corresponding region from in vitro
InvitroPartialOverlap_replacement<-bedtoolsr::bt.intersect(a=Invitro,b=PartialOverlapwithSlop,u=TRUE)

#Remove the partial overalp
Step4C<-bedtoolsr::bt.subtract(a=Step4B,b=ProteinCodingGene_Slop500,A=TRUE)

#Add the corresponding in vitro partial overlaps
Step4D<-rbind(Step4C,InvitroPartialOverlap_replacement)
```

##Step5: Merge regions too close to eachother, closer than 500bp
```{r}
Step4D_sort<-bedtoolsr::bt.sort(Step4D)
Step5<-bedtoolsr::bt.merge(i=Step4D_sort,d=500)
```

##Step6: Need to adjust peaks larger than 7000bp
```{r}
Step5_size<-Step5%>%mutate(Size=V3-V2)

#Separate ROI larger than 5000
Largerthan5000<-Step5_size%>%filter(Size>=5000) 

#Separate ROI smaller than 5000
Smallerthan5000<-Step5_size%>%filter(Size<5000)
```

Large peaks were adjusted by splitting them. For this case, there was one peak>5000bp. Which was replaced with 2 peaks from in vitro peaks.To do this, first, I got the respective peaks from in vitro. 
```{r}
#Respective peaks from in vitro
Invitro_Largerthan5000_replacement<-bedtoolsr::bt.intersect(a=Invitro,b=Largerthan5000,u=TRUE)

#Join the in vitro too large peaks with the small peaks
Step6<-rbind(Smallerthan5000%>%select(-Size),Invitro_Largerthan5000_replacement)

# readr::write_tsv(Step6,"Step6.bed.txt",col_names = F)
```

#Pos Ctrl:Add IFNG promoter
```{r}
IFNGprom<-data.frame(V1="chr12",V2=68553039,V3=68554075)

```
#All OCRs - Final target regions
```{r}
OCRs<-rbind(IFNGprom,Step6)
OCRs_sort<-bedtoolsr::bt.sort(OCRs)

```


# Neg Ctrl: CCR
So we need to find stretches of regions that are noncoding and not part of the 51 regions we are screening that can serve as negative control, regions where we can delete in the IFNG locus.

##Step1: Prepare the regions to exclude

Exclusion A: 50 candidate enhancers + 1 promoter (and 600bp buffer around them, basically 500bp and extra to be extra safe)
```{r}
ExcludeA<-bedtoolsr::bt.slop(i=OCRs_sort,g=genome_file,b=600) 
```

Exclusion B: Any gene body (and 600bp buffer around it)
```{r}
NCBI_hg19_refseq<-read.delim("SNIP-R_ScreenRegionSelection/NCBI_hg19_refseq_chr12.bed",header=F)
genome_file <-normalizePath("SNIP-R_ScreenRegionSelection/human.hg19.genome")
ExcludeB<-bedtoolsr::bt.slop(i=NCBI_hg19_refseq,g=genome_file,b=600)
# write_tsv(ExcludeB,"ExcludeB.bed.txt",col_names = F)
```

Exclusion C: Any open chromatin regions in vivo dataset
```{r}
Invivo<-read.delim("SNIP-R_ScreenRegionSelection/Yates_all_human_Tcell_ATAC.hg19.merged.bed",header=F)
ExcludeC<-bedtoolsr::bt.slop(i=Invivo,g=genome_file,b=600)
```

Combine the exclusion peaks
```{r}
Exclude_ABC<-bedtoolsr::bt.sort(rbind(ExcludeA,ExcludeB,ExcludeC))
# write_tsv(Exclude_ABC,"Exclude.bed.txt",col_names = F)
```

##Step2: Bedtools shuffle to shuffle the OCRs across the IFNG locus, avoiding the exclude peak
```{r}
# set a seed in i
Shuffle<-bedtoolsr::bt.shuffle(noOverlapping=TRUE,incl=IFNGlocus,excl=Exclude_ABC, chrom=TRUE, seed=1,i=OCRs_sort,g=genome_file)

# write_tsv(Shuffle,"shuffle.bed.txt",col_names = F)

```

##Step3: Sort and remove peaks that are closer to eachother than 600bp which will complicate gRNA design
```{r}
PotentialCCR<-bedtoolsr::bt.sort(Shuffle)
PotentialCCR_Closest<-bedtoolsr::bt.closest(t="all", d=TRUE, io=TRUE, a=PotentialCCR, b=PotentialCCR)

PotentialCCR_Closest_filtered<-PotentialCCR_Closest%>%filter(V7>600)
```

##Check: Double check to see if any are too close to protein coding gene
```{r}
Closest<-bedtoolsr::bt.closest(t="all", d=TRUE, a=PotentialCCR_Closest_filtered, b=NCBI_hg19_refseq)
```

#Export
```{r}
IFNGprom_exp<-IFNGprom%>%mutate(Type="IFNGpromoter")
Step6_exp<-Step6%>%mutate(Type="CandidateEnhancer")
CCR_exp<-PotentialCCR_Closest_filtered%>%select(one_of("V1","V2","V3"))%>%mutate(Type="CCR")
Finaldf<-rbind(IFNGprom_exp,Step6_exp,CCR_exp)
write_tsv(Finaldf,"Finaldf.bed.txt",col_names = F)
```


```{r}
sessionInfo()
```






