---
title: "02_ScreenAnalysisFinal"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(readxl)
library(tidyverse)
library(ggrepel)
library(patchwork)
```

# Common Pipeline 

## Step1: Read in files

```{r}
df<-read_excel("SNIP-R_ScreenAnalysis/SupplementaryTable4_RawScreenData.xlsx")

#separate object for each donor and rename the columns
d1<-df%>%select(Row.Barcode.IDs, ROI, gRNApairs, SimpleName, contains("d1_"))%>%rename_with(~ sub("^d1_", "", .x))
d2<-df%>%select(Row.Barcode.IDs, ROI, gRNApairs, SimpleName, contains("d2_"))%>%rename_with(~ sub("^d2_", "", .x))
d3<-df%>%select(Row.Barcode.IDs, ROI, gRNApairs, SimpleName, contains("d3_"))%>%rename_with(~ sub("^d3_", "", .x))

## Create Low bucket
d1<- d1 %>% mutate(Acute_Low=Acute_Neg+Acute_M1)
d2<- d2 %>% mutate(Acute_Low=Acute_Neg+Acute_M1)
d3<- d3 %>% mutate(Acute_Low=Acute_Neg+Acute_M1)
```

## Step2: Log normalize the counts for each column

```{r}
d1[, 5:ncol(d1)] <- lapply(d1[, 5:ncol(d1)], function(x) log2(x/sum(x)*10^6+1))
d2[, 5:ncol(d2)] <- lapply(d2[, 5:ncol(d2)], function(x) log2(x/sum(x)*10^6+1))
d3[, 5:ncol(d3)] <- lapply(d3[, 5:ncol(d3)], function(x) log2(x/sum(x)*10^6+1))
```

## Step3: Remove input not represented reads

```{r}
keep<-d1$Input_Day4 != 0 & d1$Input_Day14Acute & d2$Input_Day4 != 0 & d2$Input_Day14Acute & d3$Input_Day4 != 0 & d3$Input_Day14Acute

d1_filt<-d1[keep, ]
d2_filt<-d2[keep, ]
d3_filt<-d3[keep, ]
```

## Step4: Normalize by day 14 acute for each respective donor

```{r}
d1_filt[5:ncol(d1_filt)] <- d1_filt[5:ncol(d1_filt)] - d1_filt$Input_Day14Acute
d2_filt[5:ncol(d2_filt)] <- d2_filt[5:ncol(d2_filt)] - d2_filt$Input_Day14Acute
d3_filt[5:ncol(d3_filt)] <- d3_filt[5:ncol(d3_filt)] - d3_filt$Input_Day14Acute

```

## Step 5: Calculate Z-score for each gRNA pairs relative to NTC distribution

```{r}
Normalize_V2<-function(data,name){

# Calculate stats for normalization
Stats<-data %>%filter(SimpleName=="NonTargeting")
Mean<- Stats %>% summarise(across(where(is.numeric), mean))
Mean$Stats<- "Mean"

SD<-Stats %>% summarise(across(where(is.numeric), sd))
SD$Stats<- "sd"

#Calculate NormScore for Neg
temp <- data %>% mutate(Norm_Acute_Neg=(Acute_Neg-Mean$Acute_Neg)/SD$Acute_Neg)
temp <- temp %>% mutate(Norm_Chronic_Neg=Chronic_Neg-(Mean$Chronic_Neg)/SD$Chronic_Neg)

#Calculate NormScore for Low
temp <- temp %>% mutate(Norm_Acute_Low=(Acute_Low-Mean$Acute_Low)/SD$Acute_Low)

#Calculate NormScore for High
temp <- temp %>% mutate(Norm_Acute_High=(Acute_High-Mean$Acute_High)/SD$Acute_High)
temp <- temp %>% mutate(Norm_Chronic_High=(Chronic_High-Mean$Chronic_High)/SD$Chronic_High)

#Calculate NormScore for M2
temp <- temp %>% mutate(Norm_Acute_M2=(Acute_M2-Mean$Acute_M2)/SD$Acute_M2)
temp <- temp %>% mutate(Norm_Chronic_M2=(Chronic_M2-Mean$Chronic_M2)/SD$Chronic_M2)

#Calculate NormScore for M1
temp <- temp %>% mutate(Norm_Acute_M1=(Acute_M1-Mean$Acute_M1)/SD$Acute_M1)
temp <- temp %>% mutate(Norm_Chronic_M1=(Chronic_M1-Mean$Chronic_M1)/SD$Chronic_M1)

assign(paste("NormScore",name,sep="_"), temp,env=.GlobalEnv)

}

Normalize_V2(data=d1_filt,"d1")
Normalize_V2(data=d2_filt,"d2")
Normalize_V2(data=d3_filt,"d3")
```

# Per-gRNA pair analysis

## Step6: Center most 80% of sgRNAs for each region

```{r}
Center8<-function(df,name,OCRn,Controln){
temp<-df %>% filter(!grepl('ClosedChromatinRegion|NonTargeting',SimpleName)) %>% group_by(SimpleName) %>% mutate(AvgB4filter=mean(Norm_Acute_Low))
  temp<-temp %>% group_by(SimpleName) %>% mutate(DistanceB4filter=abs(Norm_Acute_Low-AvgB4filter))
  temp<- temp %>% group_by(SimpleName) %>% slice_min(order_by=DistanceB4filter,n=OCRn)

  ctrl<-df %>%filter(grepl('ClosedChromatinRegion|NonTargeting',SimpleName)) %>% group_by(SimpleName) %>% mutate(AvgB4filter=mean(Norm_Acute_Low))
  ctrl<-ctrl %>% group_by(SimpleName) %>% mutate(DistanceB4filter=abs(Norm_Acute_Low-AvgB4filter))
  ctrl<-ctrl %>% group_by(SimpleName) %>% slice_min(order_by=DistanceB4filter,n=Controln)
  
  out<-rbind(ctrl,temp)
  out<-out %>% select(one_of(c("ROI","SimpleName","Norm_Acute_Low","AvgB4filter","DistanceB4filter")))
  names(out)[names(out) == 'Norm_Acute_Low'] <- 'NormVal'
  assign(paste("Step6_Norm_Acute_Low",name,sep="_"), out,env=.GlobalEnv)
  
  
  
  temp<-df %>% filter(!grepl('ClosedChromatinRegion|NonTargeting',SimpleName)) %>% group_by(SimpleName) %>% mutate(AvgB4filter=mean(Norm_Acute_Neg))
  temp<-temp %>% group_by(SimpleName) %>% mutate(DistanceB4filter=abs(Norm_Acute_Neg-AvgB4filter))
  temp<- temp %>% group_by(SimpleName) %>% slice_min(order_by=DistanceB4filter,n=OCRn)
  
  ctrl<-df %>%filter(grepl('ClosedChromatinRegion|NonTargeting',SimpleName)) %>% group_by(SimpleName) %>% mutate(AvgB4filter=mean(Norm_Acute_Neg))
  ctrl<-ctrl %>% group_by(SimpleName) %>% mutate(DistanceB4filter=abs(Norm_Acute_Neg-AvgB4filter))
  ctrl<-ctrl %>% group_by(SimpleName) %>% slice_min(order_by=DistanceB4filter,n=Controln)
  
  out<-rbind(ctrl,temp)
  out<-out %>% select(one_of(c("ROI","SimpleName","Norm_Acute_Neg","AvgB4filter","DistanceB4filter")))
  names(out)[names(out) == 'Norm_Acute_Neg'] <- 'NormVal'
  assign(paste("Step6_Norm_Acute_Neg",name,sep="_"), out,env=.GlobalEnv)
}


#Check 
nrow(NormScore_d1%>%filter(SimpleName=="ClosedChromatinRegion"))
nrow(NormScore_d1%>%filter(SimpleName=="NonTargeting"))

Center8(NormScore_d1,"d1",OCRn=8,Controln=24)
Center8(NormScore_d2,"d2",OCRn=8,Controln=24)
Center8(NormScore_d3,"d3",OCRn=8,Controln=24)


```

## Plot: gRNA distribution plot of all gRNAs per donor

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

#adjust
highlight<-"0"



dataframe<-Step6_Norm_Acute_Low_d1%>%filter(SimpleName!="sgIFNG")
#run
dataframe$y<-1
line_values1<-dataframe
line_values2<-as.numeric(line_values1$NormVal)
#color
line_values1$color<-"#474F58"
line_values1$color[line_values1$SimpleName==highlight]<-"red" 
#alpha
line_values1$alpha<-"0.05"
line_values1$alpha[line_values1$SimpleName==highlight]<-"1" 
#graph
plot1<-ggplot(line_values1%>%arrange(alpha),aes(x=as.numeric(NormVal,alpha=alpha),y=y))+geom_vline(xintercept=line_values2,color=line_values1$color,alpha=as.numeric(line_values1$alpha),size=1)+ylim(c(0,0.5))+ggtitle(paste("(donor1)", highlight))+xlab("Z-score for each sgRNA pairs") +theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_alpha_continuous(range=c(0,1))+scale_x_continuous(n.breaks=6,limits=c(-4,4))


dataframe<-Step6_Norm_Acute_Low_d2%>%filter(SimpleName!="sgIFNG")
#run
dataframe$y<-1
line_values1<-dataframe
line_values2<-as.numeric(line_values1$NormVal)
#color
line_values1$color<-"#474F58"
line_values1$color[line_values1$SimpleName==highlight]<-"red" 
#alpha
line_values1$alpha<-"0.05"
line_values1$alpha[line_values1$SimpleName==highlight]<-"1" 
#graph
plot2<-ggplot(line_values1%>%arrange(alpha),aes(x=as.numeric(NormVal,alpha=alpha),y=y))+geom_vline(xintercept=line_values2,color=line_values1$color,alpha=as.numeric(line_values1$alpha),size=1)+ylim(c(0,0.5))+ggtitle(paste("(donor2)", highlight))+xlab("Z-score for each sgRNA pairs") +theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_alpha_continuous(range=c(0,1))+scale_x_continuous(n.breaks=6,limits=c(-4,4))


dataframe<-Step6_Norm_Acute_Low_d3%>%filter(SimpleName!="sgIFNG")
#run
dataframe$y<-1
line_values1<-dataframe
line_values2<-as.numeric(line_values1$NormVal)
#color
line_values1$color<-"#474F58"
line_values1$color[line_values1$SimpleName==highlight]<-"red" 
#alpha
line_values1$alpha<-"0.05"
line_values1$alpha[line_values1$SimpleName==highlight]<-"1" 
#graph
plot3<-ggplot(line_values1%>%arrange(alpha),aes(x=as.numeric(NormVal,alpha=alpha),y=y))+geom_vline(xintercept=line_values2,color=line_values1$color,alpha=as.numeric(line_values1$alpha),size=1)+ylim(c(0,0.5))+ggtitle(paste("(donor3)", highlight))+xlab("Z-score for each sgRNA pairs") +theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_alpha_continuous(range=c(0,1))+scale_x_continuous(n.breaks=6,limits=c(-4,4))


# stack vertically
(plot1 / plot2 / plot3) + plot_layout(ncol = 1)

```

# For Per-region analysis

## Step6A: Calculate per-OCR analysis

```{r}
library(dplyr)

#Select only Normalized columns that has been Z-scored

d1_PerOCR <- NormScore_d1 %>%
  group_by(SimpleName) %>%
  summarise(
    across(where(is.numeric), mean),
    .groups = "drop"
  )

d2_PerOCR <- NormScore_d2 %>%
  group_by(SimpleName) %>%
  summarise(
    across(where(is.numeric), mean),
    .groups = "drop"
  )

d3_PerOCR <- NormScore_d3 %>%
  group_by(SimpleName) %>%
  summarise(
    across(where(is.numeric), mean),
    .groups = "drop"
  )



```

## Step6B: Focusing on acute low

```{r}
d1_PerOCR_AcuteLow<-d1_PerOCR%>%select(SimpleName,Norm_Acute_Low)
d2_PerOCR_AcuteLow<-d2_PerOCR%>%select(SimpleName,Norm_Acute_Low)
d3_PerOCR_AcuteLow<-d3_PerOCR%>%select(SimpleName,Norm_Acute_Low)

d1_PerOCR_AcuteLow$donor<-"donor1"
d2_PerOCR_AcuteLow$donor<-"donor2"
d3_PerOCR_AcuteLow$donor<-"donor3"

PerOCR_AcuteLow<-rbind(d1_PerOCR_AcuteLow,d2_PerOCR_AcuteLow,d3_PerOCR_AcuteLow)

```

## Plot: gRNA distribution plot for for acute low (per OCR)

```{r, fig.width=8, fig.height=1.5}
#adjust
PerOCR_AcuteLow_Sweiss<-PerOCR_AcuteLow%>%filter(SimpleName!="sgIFNG")
highlight<-"0"

line_values1<-PerOCR_AcuteLow_Sweiss

#run
line_values1$y<-1
#color
line_values1$color<-"#474F58"
line_values1$color[line_values1$SimpleName==highlight]<-"red" 

#alpha
line_values1$alpha<-0.05
line_values1$alpha[line_values1$SimpleName==highlight]<-1 

#graph
plot3<-ggplot(line_values1%>%arrange(alpha),aes(x=as.numeric(Norm_Acute_Low,alpha=alpha),y=y)) +geom_vline(xintercept=as.numeric(line_values1$Norm_Acute_Low),color=line_values1$color,alpha=as.numeric(line_values1$alpha),size=1)+ylim(c(0,0.5))+ggtitle(highlight)+xlab("Normalized Score") +theme(axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_alpha_continuous(range=c(0,1)) +scale_x_continuous(n.breaks=10,limits=c(-2.5,2.5))
plot3

```

## Plot: X-Y concordance plot

```{r}
CP<-PerOCR_AcuteLow%>%pivot_wider(names_from = donor,values_from = Norm_Acute_Low)

CP$Color<-"OCR"
CP$Color[CP$SimpleName=="sgIFNG"]<-"sgIFNG"
CP$Color[CP$SimpleName=="0"]<-"Promoter"
CP$Color[CP$SimpleName=="ClosedChromatinRegion"]<-"ClosedChromatinRegion"
CP$Color[CP$SimpleName=="NonTargeting"]<-"NonTargeting"

palette<-c("#fdc893","#1d517b","#939598","#b16d95","#f7941d")


library(ggrepel)
Stat13<-cor.test(CP$donor1, CP$donor3, method = "pearson")

# Plot
plot1 <- ggplot(CP, aes(donor1, donor3, color = Color)) + 
  geom_point() + 
  scale_colour_manual(values = palette) +
  ggtitle("Concordance Plot") +
  theme_bw() +
  scale_alpha(range = c(0.5, 1)) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("r = ", round(Stat13$estimate, 4)),
           hjust = 3, size = 5) +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("P = ", round(Stat13$p.value, 3)),
           hjust = 1, size = 5)+xlim(c(-1.3,3.2))+ylim(c(-1.3,3.2))

plot1


Stat12<-cor.test(CP$donor1, CP$donor2, method = "pearson")
plot2 <- ggplot(CP, aes(donor1, donor2, color = Color)) + 
  geom_point() + 
  scale_colour_manual(values = palette) +
  ggtitle("Concordance Plot") +
  theme_bw() +
  scale_alpha(range = c(0.5, 1)) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("r = ", round(Stat12$estimate, 4)),
           hjust = 3, size = 5) +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("P = ", round(Stat12$p.value, 3)),
           hjust = 1, size = 5)+xlim(c(-1.3,3.2))+ylim(c(-1.3,3.2))

plot2




Stat32<-cor.test(CP$donor3, CP$donor2, method = "pearson")
plot3 <- ggplot(CP, aes(donor3, donor2, color = Color)) + 
  geom_point() + 
  scale_colour_manual(values = palette) +
  ggtitle("Concordance Plot") +
  theme_bw() +
  scale_alpha(range = c(0.5, 1)) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("r = ", round(Stat32$estimate, 4)),
           hjust = 3, size = 5) +
  annotate("text", x = 3, y = -0.9, 
           label = paste0("P = ", round(Stat32$p.value, 3)),
           hjust = 1, size = 5)+xlim(c(-1.3,3.2))+ylim(c(-1.3,3.2))


plot3

# ggsave("Figure3d_1.pdf",plot1,width=5,height=3)
# ggsave("Figure3d_2.pdf",plot2,width=5,height=3)
# ggsave("Figure3d_3.pdf",plot3,width=5,height=3)

```

# For pan-donor per-region analysis

## Step6: Join the donor data

```{r}
#Select only Normalized columns that has been Z-scored
NormScore_d1<-NormScore_d1%>%select(SimpleName,Row.Barcode.IDs,contains("Norm"))
NormScore_d2<-NormScore_d2%>%select(SimpleName,Row.Barcode.IDs,contains("Norm"))
NormScore_d3<-NormScore_d3%>%select(SimpleName,Row.Barcode.IDs,contains("Norm"))

#Rename column Normalized Value with donor number
colnames(NormScore_d1)[-c(1:2)] <- paste0("d1_", colnames(NormScore_d1)[-c(1:2)])
colnames(NormScore_d2)[-c(1:2)] <- paste0("d2_", colnames(NormScore_d2)[-c(1:2)])
colnames(NormScore_d3)[-c(1:2)] <- paste0("d3_", colnames(NormScore_d3)[-c(1:2)])

#Create a metatable
Join1<-full_join(NormScore_d1,NormScore_d2)
Join2<-full_join(Join1,NormScore_d3)

```

## Step7: Calculate pan-donor Z-score

```{r}
#Calculate Pan-donor Zscore (PDZ)
PanDonorAnalysis_AcuteLow<-Join2%>%mutate(PDZ=(d1_Norm_Acute_Low+d2_Norm_Acute_Low+d3_Norm_Acute_Low)/3)
PanDonorAnalysis_AcuteNeg<-Join2%>%mutate(PDZ=(d1_Norm_Acute_Neg+d2_Norm_Acute_Neg+d3_Norm_Acute_Neg)/3)
PanDonorAnalysis_AcuteM1<-Join2%>%mutate(PDZ=(d1_Norm_Acute_M1+d2_Norm_Acute_M1+d3_Norm_Acute_M1)/3)
PanDonorAnalysis_AcuteM2<-Join2%>%mutate(PDZ=(d1_Norm_Acute_M2+d2_Norm_Acute_M2+d3_Norm_Acute_M2)/3)
PanDonorAnalysis_AcuteHigh<-Join2%>%mutate(PDZ=(d1_Norm_Acute_High+d2_Norm_Acute_High+d3_Norm_Acute_High)/3)
```

## Step8: Calculate pan-donor normalized score

```{r}

# pull all control values once
control_pattern <- "^NonTargeting"
controls <- PanDonorAnalysis_AcuteLow %>%
  filter(grepl(control_pattern, SimpleName)) %>%
  pull(PDZ)

# for each ROI (everything except controls), compute mean + t-test vs controls
PanDonorNormalizedScore_AcuteLow <- PanDonorAnalysis_AcuteLow %>%
  filter(!grepl(control_pattern, SimpleName)) %>%   # keep only ROIs
  group_by(SimpleName) %>%
  summarise(
    meanZ = mean(PDZ, na.rm = TRUE),
    tval  = {
      x <- PDZ
      if (length(na.omit(x)) < 2 || length(na.omit(controls)) < 2) NA_real_
      else unname(t.test(x, controls)$statistic)
    },
    pval  = {
      x <- PDZ
      if (length(na.omit(x)) < 2 || length(na.omit(controls)) < 2) NA_real_
      else t.test(x, controls)$p.value
    },
    .groups = "drop"
  )

#Add NonTargeting back in just for figure purposes
controlmean<-c("NonTargeting",mean(controls),0,1)

PanDonorNormalizedScore_AcuteLow<-rbind(PanDonorNormalizedScore_AcuteLow,controlmean)

```

## Plot: Pan-donor volcano plot

```{r}

library(ggrepel)
library(dplyr)

#Set the name for each condition
PanDonorNormalizedScore_AcuteLow<-PanDonorNormalizedScore_AcuteLow%>%filter(SimpleName!="sgIFNG")
PanDonorNormalizedScore_AcuteLow$Color<-"OCR"
PanDonorNormalizedScore_AcuteLow$Color[PanDonorNormalizedScore_AcuteLow$SimpleName=="sgIFNG"]<-"sgIFNG"
PanDonorNormalizedScore_AcuteLow$Color[PanDonorNormalizedScore_AcuteLow$SimpleName=="0"]<-"Promoter"
PanDonorNormalizedScore_AcuteLow$Color[PanDonorNormalizedScore_AcuteLow$SimpleName=="ClosedChromatinRegion"]<-"ClosedChromatinRegion"
PanDonorNormalizedScore_AcuteLow$Color[PanDonorNormalizedScore_AcuteLow$SimpleName=="NonTargeting"]<-"NonTargeting"

#Set palette
palette<-c("#939598","#b16d95","#fdc893","#1d517b")


# Factor levels for consistent color + legend
PanDonorNormalizedScore_AcuteLow$Color <- factor(PanDonorNormalizedScore_AcuteLow$Color, levels = c("OCR", "sgIFNG", "Promoter", "ClosedChromatinRegion", "NonTargeting"))


# Split data into OCR vs. nonOCR
PanDonorNormalizedScore_AcuteLow_OCR <- PanDonorNormalizedScore_AcuteLow %>% filter(Color == "OCR")
PanDonorNormalizedScore_AcuteLow_rest <- PanDonorNormalizedScore_AcuteLow %>% filter(Color != "OCR")

# Plot: ROI first, then others
plot2 <- ggplot() +
  # Plot ROI first (underneath)
  geom_point(data = PanDonorNormalizedScore_AcuteLow_OCR, aes(x = as.numeric(meanZ), y = -log10(as.numeric(pval)), color = Color)) +
  
  # Plot others on top
  geom_point(data = PanDonorNormalizedScore_AcuteLow_rest, aes(x = as.numeric(meanZ), y = -log10(as.numeric(pval)), color = Color)) +
  # Label non-ROI hits
  geom_text_repel(
    data = PanDonorNormalizedScore_AcuteLow_rest %>% filter((as.numeric(pval) < 0.01) & (meanZ > 0.5)),
    aes(x = as.numeric(meanZ), y = -log10(as.numeric(pval)), label = SimpleName),
    nudge_x = 0.1,
    nudge_y = 0.1,
    colour = "black",
    max.overlaps = Inf
  ) +
  # Label OCR hits too
  geom_text_repel(
    data = PanDonorNormalizedScore_AcuteLow_OCR %>% filter((as.numeric(pval) < 0.01) & (meanZ > 0.5)),
    aes(x = as.numeric(meanZ), y = -log10(as.numeric(pval)), label = SimpleName),
    nudge_x = 0.1,
    nudge_y = 0.1,
    colour = "black",
    max.overlaps = Inf
  ) +
  scale_colour_manual(values = palette) +
  theme_bw() +
  ggtitle("IFNG acute low") +
  xlab("Pan-donor normalized score") +
  ylab("-log10(pval)") +
  scale_y_continuous(breaks = seq(0, 10, 2))+xlim(c(-2.5,2.5))
plot2

```

```{r}
sessionInfo()
```
