---
title: "01_SNIP-Rpairing"
output: html_document
date: "2026-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(readr)
```

#Step1: Prepare the peaks

[Optional] Some regions have easier time finding high scoring gRNAs if you go into the peak. So, we will employ this strategy for regions when designing gRNAs unless the region is too small to begin with. Set a cutoff. In this paper, we set the cutoff as 1000. We selected 1000bp as arbitrary cutoff as we didnt want more than 100bp/1000bp (10%) of the peak being cut into.

So first, we will clasify the region as either 500+ or <500bp
```{r}
df<-read.delim("Finaldf.bed.txt",header=F)
df<-df%>%mutate(Size=V3-V2)

#For smaller region (<=1000), no change in coordiantes
df_small<-df%>%filter(Size<=500)
#For larger region (1000+), we will set the new coordinates as (ROIstart+50,ROIend-50)
df_large<-df%>%filter(Size>500)%>%mutate(NewStart=V2+50)%>%mutate(NewEnd=V3-50)
df_large_new<-df_large%>%select(one_of(c("V1","NewStart","NewEnd","V4","Size")))
colnames(df_large_new)<-c("V1","V2","V3","V4","Size")

Step1_01<-rbind(df_large_new,df_small)
```

#Step2: Create flanks for each region
550bp flank from left and right of each region, so we will give 550bp range for gRNA selection.

```{r}
Step2_01<-Step1_01%>%mutate(Left_Start=V2-550)%>%mutate(Left_End=V2)
Step2_01<-Step2_01%>%mutate(Right_Start=V3)%>%mutate(Right_End=V3+550)
Step2_01<-Step2_01%>%dplyr::rename(Chr=V1,ROI_Start=V2,ROI_End=V3,Type=V4)

Step2_left<-Step2_01%>%select(one_of("Chr","Left_Start","Left_End"))
Step2_right<-Step2_01%>%select(one_of("Chr","Right_Start","Right_End"))

# write_tsv(Step2_left,"Step2_left.bed.txt",col_names = F)
# write_tsv(Step2_right,"Step2_right.bed.txt",col_names = F)



Step2_01$ROI<-paste(Step2_01$Chr,Step2_01$ROI_Start,Step2_01$ROI_End, sep="-")
Step2_01$LeftFlank<-paste(Step2_01$Chr,Step2_01$Left_Start,Step2_01$Left_End, sep="-")
Step2_01$RightFlank<-paste(Step2_01$Chr,Step2_01$Right_Start,Step2_01$Right_End, sep="-")

#write_tsv(Step2_01,"Step2_01_ROItoflank.bed.txt")

```
We need to separate the left and right regions and input them separately into CRISPick bc some left gRNA region of ROI overlaps with right of another ROI, and the CRISPick batch algorithm does not consider the gRNA region as indpendent and picks worse scoring guides for the second gRNA region.

#Step3: Getfasta sequence for left and right flanks

bedtools getfasta -fi chr12.fa.fai -bed CP11B_flank_left.bed -fo CP11B_output_getfasta_50bpin_left.fa
```{r}
fasta <- normalizePath("chr12.fa")
FastaLeft<-bedtoolsr::bt.getfasta(fi=fasta,bed=Step2_left)
FastaRight<-bedtoolsr::bt.getfasta(fi=fasta,bed=Step2_right)

# write_tsv(FastaLeft,"Step3_01left_fasta.txt",col_names = F)
# write_tsv(FastaRight,"Step3_01right_fasta.txt",col_names = F)
```

#Step4: Put into CRISPick (left and right separately)
Following parameters were used: Human GRCh37 (hg19), CRISPRko, SpyoCas9 (Chen tracrRNA, RuleSet3), CRISPickQuota: 20. These are suggestions, but with different CRISPick versions, the code needs adjustments. Note, due to the update on CRISPick algorithm, the output from CRISPick might be different from what was used in the paper. For the paper, CRISPick was used on June 7, 2023. For this pipeline, we reckoned that an updated CRISPick output migtht be more appropriate. Thus, CRISPick has been run on Feb 16, 2026 as an example for this pipeline. 

Should result in two txt files, one for left gRNAs, another for right gRNAs.

#Step5: gRNA pairing A
##Read in CRISPick outputs 
```{r}
Left_CRISPick<-read.delim("CRISPick/Left-sgrna-designs.txt")

Right_CRISPick<-read.delim("CRISPick/Right-sgrna-designs.txt")
```

##Step5A: Select the top 5 best sgRNAs for each flank region

This can be done by taking into account on-target and off-target score. For instance, by calculating the weighted score. Weighted score takes into account on-target and off-target ranking from crispick, where we give a slight more weight of on-target score (66%) and slight less weight of off-target score (33%). Previous versions of CRISPick output had Off target rank, but the current version do not have this. Could substitute Off target ranking with CFD based ranking. However, as recent versions of CRISPick do not output off target ranking, for the purposes of this github, I will rank the off-target based on CFD. In order to calculate the weighted score. 

```{r}

#Calculate the OffTarget Rank
Left_CRISPick_Sub <- Left_CRISPick %>%
  group_by(Input) %>%
  mutate(rank = min_rank(Aggregate.CFD.Score)) %>%
  arrange(Input, rank, Aggregate.CFD.Score) %>%
  ungroup()

Right_CRISPick_Sub <- Right_CRISPick %>%
  group_by(Input) %>%
  mutate(rank = min_rank(Aggregate.CFD.Score)) %>%
  arrange(Input, rank, Aggregate.CFD.Score) %>%
  ungroup()

Left_CRISPick_Subset<-Left_CRISPick_Sub%>%select(one_of(c("Input","sgRNA.Sequence","On.Target.Efficacy.Score","On.Target.Rank","rank")))%>%dplyr::rename(Off.Target.Rank=rank)
Right_CRISPick_Subset<-Right_CRISPick_Sub%>%select(one_of(c("Input","sgRNA.Sequence","On.Target.Efficacy.Score","On.Target.Rank","rank")))%>%dplyr::rename(Off.Target.Rank=rank)

#Set the weight
On_weight<-0.667
Off_weight<-0.333


#Calculate the weighted scores
Left_CRISPick_Subset<-Left_CRISPick_Subset %>% group_by(Input)  %>% mutate(Weighted_Score=On_weight*On.Target.Rank + Off_weight*Off.Target.Rank) #calculate the weighted score based on off target rank and on target rank
Left_CRISPick_Subset<-Left_CRISPick_Subset %>% group_by(Input)  %>% slice_min(order_by=Weighted_Score,n=5) #group by flank region > take the 5 outputs with the lowest score

Right_CRISPick_Subset<-Right_CRISPick_Subset %>% group_by(Input)  %>% mutate(Weighted_Score=On_weight*On.Target.Rank + Off_weight*Off.Target.Rank) #calculate the weighted score based on off target rank and on target rank
Right_CRISPick_Subset<-Right_CRISPick_Subset %>% group_by(Input)  %>% slice_min(order_by=Weighted_Score,n=5) #group by flank region > take the 5 outputs with the lowest score


#combine the outputs from left and right weighted score/top5 gRNAs
Step5A<-bind_rows(Left_CRISPick_Subset,Right_CRISPick_Subset)
Step5A$Input<-gsub(":","-",Step5A$Input) #replace the : to - in the chr:start-stop
names(Step5A)[names(Step5A)=="Input"]<-"flank" #rename column Input to flank
```

##Step5B: Match left and right flanks to respective ROI
```{r}
conversion<-read.delim("Step2_01_ROItoflank.bed.txt") #generated from step2
Leftflank_conversion<-conversion%>%select(ROI,LeftFlank,Type,Size)%>%dplyr::rename(flank=LeftFlank)%>%mutate(side="left")
Rightflank_conversion<-conversion%>%select(ROI,RightFlank,Type,Size)%>%dplyr::rename(flank=RightFlank)%>%mutate(side="right")
conversion_v2<-rbind(Leftflank_conversion,Rightflank_conversion)

Step5B<-full_join(Step5A,conversion_v2) #combine ROI to flank


#clean up
Step5B<- Step5B %>%arrange(ROI)%>%select(ROI,flank,sgRNA.Sequence,On.Target.Efficacy.Score)
```
#Step6: gRNA pairing A
```{r}
region_topN <- function(region,N) {
  rows <- dplyr::filter(Step5B, grepl(region,ROI)) # filter rows within all_3 that matches the ROI
   X<-split(rows, rows$flank) #split by the value in column input
 left <- as.data.frame(X[[1]]) #arbitrarily assign first as left
 right <- as.data.frame(X[[2]]) #arbitrarily assign second as right
 colnames(left) <- paste("Left", colnames(left), sep = "_") #mark left in colnames
 colnames(right) <- paste("Right", colnames(right), sep = "_") #mark right in colnames
 combbs <- tidyr::expand_grid(left,right)#pairwise comb every left row by every right row (to make this function work, need to provide separate column names)
 combbs <- dplyr::mutate(combbs, Summed_Score = Left_On.Target.Efficacy.Score + Right_On.Target.Efficacy.Score) #sum the 2 on target efficiency columns
 return(combbs %>%                                      # Top 10 highest values by group
  arrange(desc(Summed_Score)) %>% 
  dplyr::slice(1:N))
}

datalist <- lapply(unique(Step5B$ROI),region_topN,N=25) #perform the function region_topN for all the ROI then put into a dataframe # OUTPUT ALL 25!!!

Step6_final_output = do.call(rbind, datalist)
```

#Step7: Check for BSMBI cutsite
```{r}
L1<-filter(Step6_final_output, grepl("CGTCTC",Left_sgRNA.Sequence))
R1<-filter(Step6_final_output, grepl("CGTCTC",Right_sgRNA.Sequence))
L2<-filter(Step6_final_output, grepl("GAGACG",Left_sgRNA.Sequence))
R2<-filter(Step6_final_output, grepl("GAGACG",Right_sgRNA.Sequence))
```

#Step8: gRNA pairing 2

```{r}
#bring the paired gRNA that was outputed by Deb's gRNA pairing code above containing ALL 25 pairs ALREADY ordered by descending summed score


#Notes about what variables mean: max = max numbers of time to which a particular guide RNA could be outputed, region = region of interest, N = total number of guide pairs to generate. I put the max as 3 --> this means 4 unique guides will be used when 10 pairs are obtained.  

top_pairs4 <- function(region, N, max) {
  # Filter rows based on the region
  row <- dplyr::filter(Step6_final_output, grepl(region, Left_ROI))
  
  # Initialize variables
  picked_elements <- data.frame()  # Empty data frame to store entire rows
  pair_count <- 0
  
  # Initialize counters for the left and right gRNAs
  left_gRNA_count <- c()  # To store how many times each left gRNA is used
  right_gRNA_count <- c()  # To store how many times each right gRNA is used
  
  # Loop through each gRNA pair
  for (i in 1:nrow(row)) {
    pair <- row[i, ]  # Select one gRNA pair
    left_gRNA <- pair$Left_sgRNA.Sequence
    right_gRNA <- pair$Right_sgRNA.Sequence  # Extract the sequences
    
    # Count how many times the left and right gRNAs have been picked
    left_count <- sum(left_gRNA_count == left_gRNA)  # Count occurrences of left gRNA
    right_count <- sum(right_gRNA_count == right_gRNA)  # Count occurrences of right gRNA

    # Ensure that unseen gRNAs have a count of 0 (not NA)
    if (is.na(left_count)) left_count <- 0
    if (is.na(right_count)) right_count <- 0
    
    # Check if both gRNAs are under the max limit
    if (left_count < max && right_count < max) {
      # Increment the pair count and add the entire row to picked_elements
      pair_count <- pair_count + 1
      picked_elements <- rbind(picked_elements, pair)  # Add the entire pair (including ID and other columns)
      
      # Update the usage counters for left and right gRNAs
      left_gRNA_count <- c(left_gRNA_count, left_gRNA)
      right_gRNA_count <- c(right_gRNA_count, right_gRNA)
    } else {
      next  # Skip this pair if any of the sequences has been picked too many times
    }
    
    # If we've collected N pairs, break the loop
    if (pair_count == N) {
      break
    }
  }
  
  # Return the picked gRNA pairs (even if fewer than N) along with all columns
  return(picked_elements)
}

# Run the function for each unique Left_ROI
datalist4 <- lapply(unique(Step6_final_output$Left_ROI), top_pairs4, N = 10, max = 3)

datalist4<-as.data.frame(do.call(rbind,datalist4))

write_tsv(datalist4,"output.txt")

```




```{r}
sessionInfo()
```



